### 2014-09-04 introduction.tex
proof, tech, overview

\begin{figure}
\centering
\begin{tabular}{ll}
\multicolumn{2}{c}{{\tt x = 0; y = 0; z = 3;}} \\
\multicolumn{2}{c}{{\color{red} {\tt z = 2;}}} \\
\hline
\multicolumn{1}{c}{$T_1$} & \multicolumn{1}{c}{$T_2$} \\
\hline
{\tt 1: y = 3;} & \\
{\tt 2: x = 1;} & \\
{\tt 3: y = 5;} & \\
& {\tt 4: if (y > z)} \\
& {\tt 5:~~print(1/x);} \\	
& {\color{red} {\tt 6: else}} \\
& {\color{red} {\tt 7:~~print(2/x);}}
\end{tabular}
\caption{Example illustrating ordering constraints beyond synchronization primitives}
\label{fig:running}
\end{figure}

@@

 \paragraph{Our Approach} We describe a novel approach to predictive race detection, which we have implemented as the \tool\ analysis tool, that is able to relax both of these constraints. Compared to the state of the art \cite{JEFF}, \tool\ is able to detect x$1.5$ more races with only the first restriction relaxed, and x$2.5$ more races with both restrictions relaxed.

@@

 Returning to the example in Figure \ref{fig:running}, we demonstrate how value-based reasoning improves coverage. First, value-based reasoning reveals that trace 
 [{\tt 1}, {\tt 4}, {\tt 5}, {\tt 2}, {\tt 3}] is feasible, since the first assignment, {\tt y = 3}, satisfies the condition {\tt y > z}. Hence the race is discovered. A second race, between lines {\tt 2} and {\tt 7}, is detected by negating the condition. This is possible because the effects of the {\tt else} branch can be modeled precisely.
 Negation leads to an execution starting at line {\tt 4} (where {\tt y $\equiv$ 0}). The race between lines {\tt 2} and {\tt 7} then becomes visible.

>>>>

\begin{figure}
\centering
\begin{tabular}{ll}
\multicolumn{2}{c}{{\tt x = 0; y = 0;}} \\  % z = 3;
%\multicolumn{2}{c}{{\color{red} {\tt z = 2;}}} \\
\hline
\multicolumn{1}{c}{$T_1$} & \multicolumn{1}{c}{$T_2$} \\
\hline
{\tt 1: y = 3;} & \\
{\tt 2: x = 1;} & \\
{\tt 3: y = 5;} & \\
& {\tt 4: if (y > 2)} \\
& {\tt 5:~~print(1/x);} \\	
& {\color{red} {\tt 6: else}} \\
& {\color{red} {\tt 7:~~print(2/x);}}
\end{tabular}
\caption{Example illustrating ordering constraints beyond synchronization primitives}
\label{fig:running}
\end{figure}

@@

 \paragraph{Our Approach} We describe a novel approach to predictive race detection, which we have implemented as the \tool\ analysis tool, that is able to relax both of these constraints. Compared to the state of the art \cite{pldi14}, \tool\ is able to detect x$1.5$ more races with only the first restriction relaxed, and x$2.5$ more races with both restrictions relaxed.

@@

 Returning to the example in Figure \ref{fig:running}, we demonstrate how value-based reasoning improves coverage. First, value-based reasoning reveals that trace 
 [{\tt 1}, {\tt 4}, {\tt 5}, {\tt 2}, {\tt 3}] is feasible, since the first assignment, {\tt y = 3}, satisfies the condition {\tt y > 2}. Hence the race is discovered. A second race, between lines {\tt 2} and {\tt 7}, is detected by negating the condition. This is possible because the effects of the {\tt else} branch can be modeled precisely.
 Negation leads to an execution starting at line {\tt 4} (where {\tt y $\equiv$ 0}). The race between lines {\tt 2} and {\tt 7} then becomes visible.

### 2014-09-04 overview.tex
proof, tech, overview

\section{Technical Overview}\label{Se:techoverview}

@@

%TODO denote the event_{id} format?
\begin{itemize}
\item $t$ refers to the thread that issues the event, denoted as $t^e$.
\item  $id$ refers to the id associated with each event, denoted as $i^e$. The key property of  $id$ is {\em uniqueness}, i.e., any two events in the trace own different ids. The other important property of $id$ is {\em monotonicity}, i.e.,  the events from the same thread should own the strictly increasing ids. Throughout this paper, unless otherwise specified, we use  the index of an event in the trace as its id, which satisfies the above properties. Throughout this paper,  we denote an event as $e_{id}$ with the id as the subscript. We may use the terms $e_{id}$ and $id$ interchangeably given their one-to-one correspondence, e.g., $t^{e_3}$ and $t^3$. 
\item $inst$ is the static instruction. The instructions are three-address instructions involving at most three operands, which modern compilers commonly support.  Specifically, we are interested in the types of instructions listed in Table~\ref{Ta:syntax}. When the variable does not appear on the left hand of an equation, such as $y$ in $x.f=y$, it may refer to a variable, a constant or event object creation expression $new (...)$.  The $bop$ stands for the binary operator, which may refer to $+, -, *, /, \%, \wedge, \vee$ in the assignment, or refer to $<, >, =, \wedge, \vee$ in the branch. The target of the branch event is not important in our scope, therefore, we may abbreviate the branch instruction as the boolean expression afterwards. The listed instructions suffice to represent all trace events of interest. This is because a concrete finite execution trace can be reduced to a straight-line loop-free call-free path program  (argument passing and value return of the method call are modeled as assignments). This standard form of simplification preserves all the data-race-related information contained in the original trace. 
\item $map$ is a mapping from live expressions at the current event to their value (and thus a partial mapping from expressions to values). Live expressions include program variables, object fields, boolean expressions, etc. We use the notation $\lsyn exp \rsyn^e$ to retrieve the value associated with expression $exp$ at $e$.
Specially,  we also store the heap location value  $\lsyn \overline{x.f} \rsyn^e$ for a field $x.f$. Note the difference between $\lsyn \overline{x.f} \rsyn^e$ and $\lsyn x.f \rsyn^e$, which represent the location and the value stored in it respectively. The location value $\lsyn \overline{x.f} \rsyn^e$ is computed as a pair $(\lsyn x \rsyn^e, f)$.
\end{itemize}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{rcl}
			\multicolumn{1}{l}{{\tt s} $::=$} & & \\
			{\tt y = x.f} & $|$ & {\bf (heapr)} \\ 
			{\tt x.f = y}  & $|$ & {\bf (heapw)} \\ %\ $|$\ {\tt x.f = $c$}\
			{\tt z = x $bop$ y}\  & $|$ & {\bf (assign)} \\ %$|$\ {\tt z = $c$} $|$ {\tt z = new()}
			{\tt if (x $bop$  y) goto ...} & $|$ &  {\bf (branch)} \\
			{\tt lock(l)}\ $|$\ {\tt unlock(l)}  & $|$& {\bf (sync)} \\
			{\tt fork(t)}\ $|$\ {\tt join(t)}$|$ {\tt begin(t)}\ $|$\ {\tt end(t)} &  & {\bf (thread)}
		\end{tabular}
	\end{center}
	\caption{\label{Ta:syntax}Language syntax}
\end{table}

{\bf Trace \ } 
We propose the projection operation and domains to facilitate the reasoning of the trace.
\begin{itemize}
\item projection operation  ``$|$'': $\tau|t$ contains only the (ordered) events from the thread $t$; $\tau|l$ contains the events involving the location $l$; $\tau|i$ contains only the event of which the id equals $i$; $\tau|\leq e$ denotes the prefix of $e$, i.e., the events preceding the event $e$; $\tau|\geq e$ denotes the events after $e$; $\tau|>=e_1\wedge <=e_2$ denotes the events between $e_1$ and $e_2$.
\item sets: $\mathcal{T}$ denotes the set of threads involved;  $\mathcal{L}$ denotes the set of memory locations involved; 
$\mathcal{SV}$ denotes the set of shared variables, which reference the shared locations, the shared location $l$ can be judged by counting the number of threads in $\tau|l$;  $\mathcal{E}$ denotes the set of events involved.
\end{itemize}

@@

Given the trace, the predictive analysis preserves the same event sequence for each thread, i.e., $T_1$ still executes $e_1 e_2 e_3$ and $T_2$ still executes $e_4 e_5$, 
while rescheduling the events from different threads so that the potential race pair can run concurrently, i.e., $e_2$ and $e_5$ run concurrently. Therefore, our goal is to compute a new schedule that witnesses a race, under the (both inter-thread and intra-thread) data flow constraints, control flow constraints and synchronization constraints. In the following, we explain only the necessary constraints (the full description of constraints is explained in Section~\ref{sec:tech}).

@@

To model the schedule, we assign an order (integer) variable to each event, e.g., $O_{e_1}$. The event with smaller order variable is scheduled earlier.
Also, we introduce symbols to denote shared accesses, e.g., $W^{id}_{x}$ ($R^{id}_{x}$) denotes the write (read) of the shared variable $x$ by the event $e_{id}$.
The local accesses are not used in this example and therefore omitted.

@@

{\bf Race Condition\ } The potential race pair involves two accesses of the same shared variable from different threads, where one access is a write. Given any potential race pair, e.g., ($e_2$, $e_5$), it is a real race if and only if the two accesses can occur at the same time (race condition), which is captured by the race condition constraint
$$
	O_2 = O_5
$$.
The race condition constraint--- together with the other constraints --- guarantees the feasibility of the predicted race if a solution is found for the overall constraint system.

@@

\item {\bf Intra-thread Data Flow Constraints\ } The events should respect the data flow constraints imposed by each instruction.
 For our running example, we obtain:
$$
	W_{\tt x}^0 = 0 \wedge W_{\tt y}^0 = 0 \wedge W_{\tt z}^0 = 3 \wedge W_{\tt y}^1 = 3 \wedge
			W_{\tt x}^2 = 1 \wedge W_{\tt y}^3 = 5
$$
As an example, $W_{\tt y}^3= 5$ denotes that the value assigned to variable {\tt y}  by event $e_3$ is $5$.
\end{itemize}

@@

Indeed, this constraint is satisfied by the assignments to {\tt y} both at line {\tt 1} and at line {\tt 3}.

@@

{\bf Unexplored Branches\ } Beyond the encoding steps so far, which focus on the given trace, we can often encode constraints along unexplored branches. In our running example, this is essential to discover the race between lines {\tt 2} and {\tt 7}.

The conflicting accesses in this case, determined based on static analysis of $P$, are expressed as $O_2=O_7$. In addition, we negate the path condition, thereby obtaining $R_{\tt y}^4 > 2 \equiv {\bf false}$ in place of $R_{\tt y}^4 > 2 \equiv {\bf true}$. We also model the other necessary constraints such as the intra-thread order constraint $O_4 < O_7$.

{\bf Constraint Solving\ } Having conjoined the formuals from the different encoding steps into (i) a global representation of all feasibility constraints (path, ordering, assignment and other constraints) and (ii) the requirement for a given race to occur (expressed as simultaneous execution of the conflicting accesses), we discharge the resulting formula to an off-the-shelf constraint solver, such as Z3 or Yices. If successful, the solver returns a solution for the specified constraints.

In particular, the solution discloses a feasible trace that gives rise to the race at hand. The trace is identified uniquely via the order enforced in the solution over the variables $O_i$, which represent scheduling order.

>>>>

\section{Technical Overview}\label{sec:overview}

@@

%TODO denote the event_{id} format?
\begin{itemize}
\item $t$ refers to the thread that issues the event, denoted as $t^e$.
\item  $id$ refers to the id associated with each event. The key property of  $id$ is {\em uniqueness}, i.e., any two events in the trace own different ids. The other important property of $id$ is {\em monotonicity}, i.e.,  the events from the same thread should own the strictly increasing ids. Throughout this paper, unless otherwise specified, we use  the index of an event in the trace as its id, which satisfies the above properties. Throughout this paper,  we denote an event as $e_{id}$ with the id as the subscript. We may use the terms $e_{id}$ and $id$ interchangeably given their one-to-one correspondence, e.g., $t^{e_3}$ and $t^3$. 
\item $inst$ is the static instruction. The instructions are three-address instructions involving at most three operands, which modern compilers commonly support.  Specifically, we are interested in the types of instructions listed in Table~\ref{Ta:syntax}. When the variable does not appear on the left hand of an equation, such as $y$ in $x.f=y$, it may refer to a variable, a constant or event object creation expression $new (...)$.  The $bop$ stands for the binary operator, which may refer to $+, -, *, /, \%, \wedge, \vee$ in the assignment, or refer to $<, >, =, \wedge, \vee$ in the branch. The target of the branch event is not important in our scope, therefore, we may abbreviate the branch instruction as the boolean expression afterwards. The listed instructions suffice to represent all trace events of interest. This is because a concrete finite execution trace can be reduced to a straight-line loop-free call-free path program  (argument passing of the method call is modeled as assignments and the virtual call resolution is modeled as branches). This standard form of simplification preserves all the data-race-related information contained in the original trace. 
\item $map$ is a mapping from live expressions at the current event to their value (and thus a partial mapping from expressions to values). Live expressions include program variables, object fields, boolean expressions, etc. We use the notation $\lsyn exp \rsyn^e$ to retrieve the value associated with expression $exp$ at $e$.
Specially,  we also store the heap location value  $\lsyn \overline{x.f} \rsyn^e$ for a field $x.f$. Note the difference between $\lsyn \overline{x.f} \rsyn^e$ and $\lsyn x.f \rsyn^e$, which represent the location and the value stored in it respectively. The location value $\lsyn \overline{x.f} \rsyn^e$ is computed as a pair $(\lsyn x \rsyn^e, f)$.
\end{itemize}

@@

\begin{table}
	\begin{center}
		\begin{tabular}{rcl}
			\multicolumn{1}{l}{{\tt s} $::=$} & & \\
			{\tt y = x.f} & $|$ & {\bf (heapr)} \\ 
			{\tt x.f = y}  & $|$ & {\bf (heapw)} \\ %\ $|$\ {\tt x.f = $c$}\
			{\tt z = x $bop$ y}\  & $|$ & {\bf (assign)} \\ %$|$\ {\tt z = $c$} $|$ {\tt z = new()}
			{\tt if (x $bop$  y) goto ...} & $|$ &  {\bf (branch)} \\
			{\tt lock(l)}\ $|$\ {\tt unlock(l)}  & $|$& {\bf (sync)} \\
			{\tt fork(t)}\ $|$\ {\tt join(t)}$|$ {\tt begin(t)}\ $|$\ {\tt end(t)} &  & {\bf (thread)}
		\end{tabular}
	\end{center}
	\caption{\label{Ta:syntax}Instruction Types}
\end{table}

{\bf Trace \ } 
We introduce the projection operation ``$\downarrow$'' to facilitate the reasoning of the trace:
$\tau\downarrow_{t}$ contains only the (ordered) events from the thread $t$; $\tau\downarrow_{\ell}$ contains the events involving the location $\ell$; $\tau\downarrow_{\leq id}$ denotes the prefix of $e_{id}$, i.e., the events preceding the event $e_{id}$; $\tau\downarrow_{\geq id}$ denotes the events after $e_{id}$; $\tau\downarrow_{>=id1\wedge <=id2}$ denotes the events between $e_{id1}$ and $e_{id2}$.

%\begin{itemize}
%\item projection operation  : 
%%\item sets: $\mathcal{T}$ denotes the set of threads involved;  $\mathcal{L}$ denotes the set of memory locations involved; 
%%$\mathcal{SV}$ denotes the set of shared variables, which reference the shared locations, the shared location $l$ can be judged by counting the number of threads in $\tau\downarrow_{\ell}$;  $\mathcal{E}$ denotes the set of events involved.
%\end{itemize}

@@

Given the trace, the predictive analysis preserves the same event sequence for each thread, i.e., $T_1$ still executes $e_1 e_2 e_3$ and $T_2$ still executes $e_4 e_5$, 
while rescheduling the events from different threads so that the potential race pair can run concurrently, i.e., $e_2$ and $e_5$ run concurrently. Therefore, our goal is to compute a new schedule that witnesses a race, under the (both inter-thread and intra-thread) data flow constraints, control flow constraints and synchronization constraints. In the following, we explain only the necessary constraints (the full description of constraints is explained in Section~\ref{sec:relax1}).

@@

To model the schedule, we assign an (integer) order  variable to each event, e.g., $O_{e_1}$. The event with smaller order variable is scheduled earlier.
Also, we introduce symbols to denote shared accesses, e.g., $W^{id}_{x}$/$R^{id}_{x}$ denotes the write/read of the shared variable $x$ by the event $e_{id}$.
The local accesses are not used in this example and therefore omitted.

@@

{\bf Race Condition\ } The potential race pair involves two accesses of the same shared location from different threads, where at least one access is a write. Given any potential race pair, e.g., ($e_2$, $e_5$), it is a real race if and only if the two accesses can occur at the same time (race condition), which is captured by the race condition constraint
$$
	O_2 = O_5
$$.
The race condition constraint--- together with the other constraints --- guarantees the feasibility of the predicted race if a solution is found for the overall constraint system.

@@

\item {\bf Intra-thread Data Flow Constraints\ } The events should respect the data flow constraints imposed by each instruction.
 For our running example, we obtain:
$$
	W_{\tt x}^0 = 0 \wedge W_{\tt y}^0 = 0 \wedge W_{\tt y}^1 = 3 \wedge
			W_{\tt x}^2 = 1 \wedge W_{\tt y}^3 = 5
$$
As an example, $W_{\tt y}^3= 5$ denotes that the value assigned to variable {\tt y}  by event $e_3$ is $5$.
\end{itemize}

@@

%Indeed, this constraint is satisfied by the assignments to {\tt y} both at line {\tt 1} and at line {\tt 3}.

@@

{\bf Unexplored Branches\ } Beyond the encoding steps so far, which focus on the given trace, we can often encode constraints along unexplored branches. In our running example, this is essential to discover the race between lines {\tt 2} and {\tt 7}. We conduct the symbolic execution to exercise the unexplored branches, which returns a set of traces representing possible executions of the unexplored branch. Each trace is combined with the original trace to find races involving the accesses in the unexplored path. For the racy events, $e_2$ and $e_7$, we first specify the race condition constraint  $O_2=O_7$. In addition, we negate the path condition, thereby obtaining $R_{\tt y}^4 > 2 \equiv {\bf false}$ in place of $R_{\tt y}^4 > 2 \equiv {\bf true}$. We also model the other necessary constraints such as the intra-thread order constraint $O_4 < O_7$, similar to the above analysis.

{\bf Constraint Solving\ } The formuals from the different encoding steps are conjoined and sent to the off-the-shelf solver such as {\sf z3}.  The race under analysis is real if the solver returns a solution, which includes the scheduling order and the values for the variables.

{\bf Highlights \ } In this example, our analysis finds two more pairs of races than existing analyses~\cite{yannis, jeff}, which include $(e_2, e_5)$ and $(e_2, e_7)$.
The key insight for finding the first race is, we do not require $e_4$ to read from $e_3$ to retain the original value, as what existing approaches do, rather, we allow $e_4$ to read from $e_1$ while still preserving the feasibility of the branch. This relaxation allows the events after $e_4$ to run concurrently with the events before $e_3$, leading to the detection of the first race.  The key insight for finding the second race is, we explore the unexplored branches and ask the solver to compute a feasible schedule that exercises them. The neighboring branches are likely to contain racy events if the executed branches contain racy events.

%In particular, the solution discloses a feasible trace that gives rise to the race at hand. The trace is identified uniquely via the order enforced in the solution over the variables $O_i$, which represent scheduling order.

### 2014-09-04 relax2.tex
proof, tech, overview

\section{Exploration of Unexecuted Branches}
We now switch to the second feature of \tool, which is its ability to reason about unexplored branches.
At the high level, we simply need to derive a trace that contains the unexplored branch, after that, we can reduce the analysis to the analysis in Section~\ref{sec:relax} (starting from the SSA processing).  \tool\ derives such traces by symbolically executing the unexplored branches to record the events and including these events in the new traces.

@@

\begin{algorithmic}[3]
\For {$e: \tau$}
  \If {$type(e)=branch$}
    \State $S=symEngine(\tau, e)$
	\For {$\tau': S$}
	\State $\tau_r=\tau-((\tau|t^e)|>=e)$
	\State $\tau_r=\tau_r.replace(e, neg(e) + \tau')$
	\EndFor
 \EndIf
\EndFor
\end{algorithmic}

@@

The analysis starts with the event $e$ (exclusively), with the current state $\sigma$ at $e$ as the initial state. 
Remember that we use the runtime values for two types of events: the heap value of the base objects and the evaluation result of the boolean expression in the branch. 
Other values are maintained but not used. Therefore, we also need to maintain the heap value and the boolean expression value in the symbolic execution.

First, we propose standard kill/gen rules for three basic heap instructions to maintain the heap update along a given path. Specially, we separate the object creation from the local assignment. In the object creation, we assign a unique integer to represent the newly created object and store it into the state $\sigma$. For field reference, $\sigma[x.f]$ is actually short for two-level interpretation $\sigma[\sigma[x].f]$, where the first finds out the base object $\sigma[x]$ and the second searches for the object referenced by its field $f$. The kill/gen rules take effect only when $x.f$ is not of the primitive type.

>>>>

\section{Exploration of Unexecuted Branches}~\label{sec:relax2}
We now switch to the second feature of \tool, which is its ability to reason about unexplored branches.
At the high level, we simply need to derive a trace that contains the unexplored branch, after that, we can reduce the analysis to the analysis in Section~\ref{sec:relax} (starting from the SSA processing).  \tool\ derives such traces by symbolically executing the unexplored branches to record the events and including these events in the new traces.

@@

\begin{algorithmic}[3]
\For {$e: \tau$}
  \If {$type(e)=branch$}
    \State $S=symEngine(\tau, e)$
	\For {$\tau': S$}
	\State $\tau_r=\tau-\tau\downarrow_{t^e \wedge \geq e}$
	\State $\tau_r=\tau_r.replace(e, neg(e) + \tau')$
	\EndFor
 \EndIf
\EndFor
\end{algorithmic}

@@

The analysis starts at the event $e$ (exclusively), with the state $\sigma$.  Initially, the state includes the runtime heap value for the base objects that are already resolved. During the symbolic execution, the state is updated to maintain the heap value and also the boolean expression values for branches. The values for other variables are not used indeed and therefore treated as symbolic. The update of the state is realized through the   standard kill/gen rules for three basic heap instructions. In the object creation, which we separate from the local assignment, we assign a unique integer to represent the newly created object and store it into the state $\sigma$. For field access, $\sigma[x.f]$ actually denotes $\sigma[o.f]$, where  $o$ is the base object resolved in the state. 
These kill/gen rules take effect only when $x.f$  is not of the primitive type.
